%% ============================================================
%%  CHAPTER 1 — INTRODUCTION
%% ============================================================
\chapter{Introduction}
\label{ch:intro}

\section{Motivation and Background}
\label{sec:motivation}

The global transition towards electrified transportation and grid-level energy storage has placed lithium-ion battery technology at the centre of modern engineering research. Electric vehicles (EVs), stationary storage systems, and consumer electronics collectively account for demand that is projected to exceed 2,000~\si{\giga\watt\hour} annually by 2030~\cite{IEA2023}. Ensuring the safety, reliability, and longevity of these systems requires a quantitative, physics-based understanding of the electrochemical, thermal, and mechanical processes occurring within and between individual battery cells.

A modern lithium-ion battery pack is a deeply heterogeneous system. Even when manufactured with nominally identical cells, small differences in production tolerances, contact resistance at bus bars, thermal positioning within the pack enclosure, and the evolution of the Solid Electrolyte Interphase (SEI) layer over successive charge--discharge cycles combine to produce significant spread in cell-level state-of-charge (SOC), temperature, and state-of-health (SOH). This heterogeneity cascades: a hotter cell ages faster, its increased internal resistance forces neighbouring cells to carry a greater share of the pack current, further accelerating the thermal imbalance---a self-reinforcing feedback that ultimately governs pack lifetime~\cite{Gogoana2014,Baumhofer2014}.

Understanding this feedback demands models that simultaneously resolve:
\begin{enumerate}[noitemsep]
  \item the micro-scale electrochemical kinetics within solid electrode particles (nanometre to micrometre);
  \item the mesoscale electrolyte transport across electrode and separator layers (micrometre to millimetre);
  \item the cell-level thermal dynamics (centimetre scale);
  \item the pack-level current sharing and thermal conduction between cells (decimetre to metre scale).
\end{enumerate}

The Doyle--Fuller--Newman (DFN) model~\cite{Doyle1993,Fuller1994} achieves fidelity at the cell level by solving coupled PDEs for solid and electrolyte phase concentrations and potentials. However, the stiffness of these equations---arising from the wide separation of diffusive timescales between the solid particle (characteristic time $R_s^2/D_s \sim 10^3$~s) and the electrolyte (characteristic time $L^2/D_\text{eff} \sim 10$~s)---makes direct pack-level DFN simulation computationally prohibitive without specialised numerical methods.

\subsection{The Computational Challenge}
\label{ssec:comp_challenge}

The typical DFN model for a single cell involves $O(100)$ coupled unknowns per time step and a dense Jacobian. Extending this naively to a $N_\text{s} \times N_\text{p}$ pack of $N_\text{cells} = N_\text{s} \cdot N_\text{p}$ cells produces a system of size $O(100 N_\text{cells})$, with cross-cell coupling through both the electrical network and the thermal conduction matrix. The Jacobian of the full coupled system scales as $O(N_\text{cells}^2)$ in the worst case, making direct dense methods infeasible beyond a few dozen cells.

Contemporary literature reports that:
\begin{itemize}[noitemsep]
  \item DFN simulations of even 12-cell packs are described as computationally intensive~\cite{Kim2011};
  \item Most pack-level electrochemical studies are restricted to series strings of 4--10 cells~\cite{Legrand2014};
  \item Pack-level degradation studies almost universally employ reduced-order models such as the Single Particle Model (SPM) or equivalent-circuit models (ECM)~\cite{Zhang2020,Plett2004}.
\end{itemize}

The present work demonstrates that these computational barriers can be overcome through:
\begin{itemize}[noitemsep]
  \item a \emph{block-diagonal} Jacobian structure, exploiting the independence of per-cell physics while resolving pack-level coupling efficiently;
  \item GPU-accelerated automatic differentiation via \texttt{torch.vmap} and \texttt{torch.func.jacrev}, computing exact gradients simultaneously across the pack;
  \item an innovative implicit Backward Euler solver strategy with scaled Newton iterations designed specifically for PyTorch compatibility and massively parallel execution.
\end{itemize}

\section{Literature Gap and Research Motivation}
\label{sec:literature_gap}

\Cref{tab:pack_comparison} summarises representative pack-level battery simulation studies from the literature. The pattern is striking: no existing open framework simultaneously provides DFN-level electrochemical fidelity, coupled thermal modelling, mechanistic degradation, large-scale parallelism, and modular software extensibility in a single integrated tool.

\begin{table}[H]
\centering
\caption{Comparison of representative battery pack simulation frameworks in the literature.}
\label{tab:pack_comparison}
\small
\begin{tabularx}{\textwidth}{p{3.0cm}ccccccc}
\toprule
\textbf{Framework / Study} & \textbf{Max Cells} & \textbf{Thermal} & \textbf{Degradation} & \textbf{Electrochem.} & \textbf{GPU Fast} & \textbf{Topology} \\
\midrule
Doyle et al.\ \cite{Doyle1993}      & 1   & No  & No  & DFN  & No  & Single \\
Kim et al.\ \cite{Kim2011}          & 12  & Yes & No  & DFN  & No  & Series \\
Legrand et al.\ \cite{Legrand2014}  & 6   & No  & No  & SPMe & No  & Series \\
PyBaMM \cite{Sulzer2021}            & 1   & Yes & Yes & DFN  & No  & Single \\
Zhang et al.\ \cite{Zhang2020}      & 100 & Yes & ECM & ECM  & No  & S/P Pack \\
Gogoana et al.\ \cite{Gogoana2014}  & 25  & No  & Est & ECM  & No  & Parallel \\
\textbf{\VIBE{} (this work)}        & \textbf{400+} & \textbf{Yes} & \textbf{Yes} & \textbf{DFN} & \textbf{Yes} & \textbf{Arbitrary} \\
\bottomrule
\end{tabularx}
\end{table}

While many papers have developed pack-level physics-based models, most are limited to specific purposes such as heterogeneity analysis or balancing strategies, and lack speed because they do not utilize GPU acceleration. Consequently, they cannot be expanded for larger packs and are often restricted to only series or only parallel configurations. This work bridges that gap by providing a fully coupled, fast, and scalable pack-level modeling framework.

\section{Objectives of this Thesis}
\label{sec:objectives}

The primary objective of this thesis is the design, implementation, and validation of a modular, GPU-accelerated software framework for coupled electrochemical--thermal--degradation simulation of large-scale heterogeneous lithium-ion battery packs. The specific research objectives are:

\begin{enumerate}
  \item \textbf{Implement a GPU-accelerated scalable pack framework:} Develop an innovative solver strategy that explicitly structures the cell-state matrices for PyTorch compatibility. By resolving the exact math of batched Jacobians, the framework overcomes the traditional computational bottlenecks, allowing large arbitrary $N_\text{s} \times N_\text{p}$ topologies to be simulated efficiently on GPUs.
  
  \item \textbf{Integrate high-fidelity physics models from literature:} Incorporate established electrochemical, thermal, and SEI layer equations referenced from well-known literature (such as PyBaMM). The novelty lies not in new equations, but in the fast, robust integration of everything (stress, SEI, thermal) with the capability to easily add more detail.

  \item \textbf{Establish a highly modular, plug-and-play architecture:} Design a framework where thermal models, SEI mechanisms, and control strategies can be changed or made more detailed easily without touching core solver mechanics. Block diagrams and workflows will demonstrate the extreme flexibility of the framework.

  \item \textbf{Construct a fully coupled pack-level model:} Model detailed pack-level current sharing, thermal interconnection, and heterogeneity propagation simultaneously without sacrificing execution speed, addressing the limitations of existing CPU-bound studies.

  \item \textbf{Demonstrate computational scaling and capabilities:} Conduct single-cell and SEI model comparisons with PyBaMM, perform speed comparisons between CPU and GPU to demonstrate how the exponential scaling barrier is broken, and simulate severe pack heterogeneity (e.g., $10s10p$ uneven cooling) to visualize coupled uneven SEI growth and current variations.
  
  \item \textbf{Conduct fast, exact local sensitivity analysis:} Highlight how the fully differentiable framework makes complex sensitivity analysis significantly faster and easier compared to literature baselines, facilitating automated gradient computation for cell variations.
  
  \item \textbf{Provide general numerical flexibility as a secondary benefit:} While the focus is heavily on system integration and GPU performance, the framework will still allow operators (like FVM) to be customized easily.
\end{enumerate}

\section{Thesis Contributions}
\label{sec:contributions}

The original contributions of this thesis are as follows:

\begin{description}
  \item[C1 --- Fast, GPU-Batched Pack Solver for PyTorch.] An innovative implicit Newton solver designed specifically for PyTorch. By structuring the physics into large batched matrices and utilizing \texttt{torch.vmap} with \texttt{jacrev}, the solver computes the fully coupled PDE system in parallel, achieving near-linear GPU scaling with pack size and breaking the conventional bottleneck of exponential scaling beyond 20 cells.
  
  \item[C2 --- Deeply Coupled Multipack Physics Integration.] While utilizing standard literature equations for individual cell chemistry, SEI, and thermal dynamics (referenced from PyBaMM and related works), this work achieves the unprecedented integration of full DFN-class electrochemical kinetics, thermal interconnection, and differential SEI degradation simultaneously at the pack level.

  \item[C3 --- Plug-and-Play Modular Architecture.] A highly flexible, modular registry where new control strategies (charging, cooling) and degradation models (seven distinct SEI models) can be swapped seamlessly. Block diagrams illustrate the ease with which different strategies from the literature can be implemented without touching the underlying core solver logic.

  \item[C4 --- Exact Local Sensitivity Analysis via Automatic Differentiation.] By leveraging the fully differentiable tensor architecture, the framework performs exact local sensitivity gradients (e.g., $\partial \dot{T} / \partial p_i$) of arbitrary model outputs. This capability drastically simplifies and accelerates sensitivity studies relative to existing perturbation methods in the literature.
  
  \item[C5 --- Generalised PDE Operator Support.] Although not the primary focus, the framework still provides a declarative operator-based pipeline that supports generic discretisation methods (like FVM for exact flux conservation), serving as a robust mathematical foundation for the high-performance solver.
\end{description}

\section{Thesis Organisation}
\label{sec:organisation}

The remainder of this thesis is organised as follows:

\begin{description}[leftmargin=2em]
  \item[Chapter~\ref{ch:literature}] reviews the literature on electrochemical battery models, numerical methods for battery simulation, degradation physics, and pack-level studies.

  \item[Chapter~\ref{ch:math_model}] derives the complete mathematical model: solid diffusion, electrolyte transport, Butler--Volmer kinetics, OCP parameterisation, thermal model, and the pack-level electrical network.

  \item[Chapter~\ref{ch:numerics}] presents the numerical methods: Chebyshev collocation, finite volume discretisation, the Newton--Raphson solver, Jacobian scaling, adaptive timestepping, and GPU batching.

  \item[Chapter~\ref{ch:software}] describes the software architecture: the symbolic PDE AST, operator pipeline, state layout registry, and the modular physics API.

  \item[Chapter~\ref{ch:degradation}] details the degradation physics: SEI growth mechanisms, lithium plating, mechanical stress, and crack-enhanced SEI formation.

  \item[Chapter~\ref{ch:pack}] presents the pack-level simulation framework: current distribution, thermal interconnection, and heterogeneity modelling.

  \item[Chapter~\ref{ch:control}] covers battery management: balancing strategies and thermal management modules.

  \item[Chapter~\ref{ch:results}] presents validation and simulation results.

  \item[Chapter~\ref{ch:conclusion}] concludes the thesis and outlines future research directions.
\end{description}
